---
title: Build a data frame from vectors
runtime: shiny_prerendered
tutorial:
  version: 0.1
  id: course-r-introduction#02-data-frames-tibbles#01-build-data-frame-from-vectors
output:
  learnr::tutorial:
    df_print: default
    progressive: false
    allow_skip: false
usagePlan: free
---

```{r setup, include=FALSE}
library(testwhat)
library(learnr)

tutorial_options(exercise.timelimit = 60, exercise.checker = testwhat_learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## Build a data frame from vectors

Tabular data is the most common format used by data scientists. In R, tables are represented through data frames. They can be inspected by printing them to the console.

- Understand why data frames are important
- Interpret console output created by a data frame
- Create a new data frame using the `data.frame()` function
- Define vectors to be used for single columns
- Specify names of data frame columns

```{r eval=FALSE, recipe=TRUE}
data.frame(___ = ___, 
           ___ = ___, 
           ...)
```

## Introduction to Data Frames

In analysis and statistics, tabular data is the most important data structure. It is present in many common formats like Excel files, comma separated values (CSV) or databases. R integrates tabular data objects as first-class citizens into the language through *data frames*. Data frames allow users to easily read and manipulate tabular data within the R language.

Let's take a look at a data frame object named `Davis`, from the package **carData**, which includs height and weight measurements for 200 men and women:

```{r, include=FALSE}
library(carData)
options(max.print = length(Davis) * 3)
Davis
```

```{r, echo = TRUE}
Davis
```

From the printed output we can see that the data frame spans over 200 **rows** (3 printed, 197 omitted) and 5 **columns**. In the example above, each row contains data of one person through **attributes**, which correspond to the columns `sex`, `weight`, `height`, reported weight `repwt` and reported height `repht`. 

For example, the first row in the table specifies a `M`ale weighing `77`kg and has a height of `182`cm. The reported weights are very close with `77`kg and `180`cm, respectively.

The rows in a data frame are further identified by *row names* on the left which are simply the row numbers by default. In the case of the `Davis` dataset above the row names range from 1 to 200.

## Quiz: Data Frame Output

<!--does the output look the same on the website? choose other dataset?-->
```{r, echo=FALSE}
library(carData)
options(max.print = length(Salaries) * 3)
Salaries
```

The data frame above shows the nine-month academic salary for Assistant Professors, Associate Professors and Professors in a college in the U.S.

```{r quiz-02-01-01, echo = FALSE}
question('Which answers about the data frame printed above are correct?',
  answer("The data frame has 3 rows.", message = "There have been some rows omitted at the data frame output as indicated by the bottom line."),
  answer("The data frame has 394 rows.", message = "The bottom line of the data frame output shows only the omitted rows."),
  answer("The data frame has 397 rows.", correct = TRUE),
  answer("The data frame has 6 attributes.", correct = TRUE),
  answer("The attribute names contain `Prof` and `AsstProf`", message = "These are the attribute values of `rank`."),
  allow_retry = TRUE
)
```

## Quiz: Data Frame Output (2)

<!--does the output look the same on the website? choose other dataset?-->
```{r, echo=FALSE}
library(carData)
options(max.print = length(Salaries) * 3)
Salaries
```

The data frame above shows the nine-month academic salary for Assistant Professors, Associate Professors and Professors in a college in the U.S.

```{r quiz-02-01-02, echo = FALSE}
question('Which answers about the first three faculty members are correct?',
  answer("All three are male.", correct = TRUE),
  answer("The salaries of all three members are about the same.", message = "The two `Prof`essors in the first two rows earn about double compared to the Assistant Professor in row three."),
  answer("The Professor in row three is most probably be the oldest.", message = "The professor in row three most probably the youngest according to `yrs.since.phd` and `yrs.service`."),
  answer("All shown professors are from the same discipline.", correct = TRUE),
  answer("The highest salary amongst the three Professors is $139,750.", message = "The highest salary is $173,200."),
  allow_retry = TRUE
)
```

## Creating Data Frames

```{r eval=FALSE, highlightLines=1:3}
data.frame(___ = ___, 
           ___ = ___, 
           ...)
```

Data frames hold tabular data in various columns or *attributes*. Each column is represented by a vector of different *data types* like numbers or characters. The `data.frame()` function supports the construction of data frame objects by combining different vectors to a table. To form a table, vectors are required to have equal lengths. A data frame can also be seen as a collection of vectors connected together to form a table.

Let's create our first data frame with four different people including their ids, names and indicators if they are female or not. Each of these attributes is created by a different vector of different data types (numeric, character and logical). The attributes are finally combined to a table using the `data.frame()` function:

```{r, echo=TRUE}
data.frame(
  c(1, 2, 3, 4),
  c("Louisa", "Jonathan", "Luigi", "Rachel"),
  c(TRUE, FALSE, FALSE, TRUE)
)
```

The resulting data frame stores the values of each vector in a different column. It has four rows and three columns. However, the column names printed on the first line seem to include the column values separated by dots which is a very strange naming scheme! 

Column names can be included into the `data.frame()` construction as argument names preceding the values of column vectors. To improve the column naming of the previous data frame we can write

```{r, echo=TRUE}
data.frame(
  id = c(1, 2, 3, 4),
  name = c("Louisa", "Jonathan", "Luigi", "Rachel"),
  female = c(TRUE, FALSE, FALSE, TRUE)
)
```

The resulting data frame includes the column names needed to see the actual meaning of the different columns.

## Exercise: Creating Your First Data Frame

```{r, results='asis', echo=FALSE}
knitr::kable(data.frame(
  weekday = c("Monday", "Tuesday", "Wednesday"),
  temperature = c(28, 31, 25),
  hot = c(FALSE, TRUE, FALSE)
))
```

Let's create a data frame as shown above using the `data.frame()` function. The resulting data frame should consist of the three columns `weekday`, `temperature` and `hot`:

1. The first column named `weekday` contains the weekday names `"Monday"`, `"Tuesday"`, `"Wednesday"`.
2. The second column named `temperature` contains the temperatures (in degrees Celsius) as `28`, `31`, `25`.
3. The third column named `hot` contains the logical values `FALSE`, `TRUE`, `FALSE`.

Store the final data frame in the variable `temp` and print its output to the console:

```{r exercise-02-01-03, exercise=TRUE}
temp <- data.frame(
  

)
temp
```

```{r exercise-02-01-03-hint-1, eval=FALSE}
temp <- data.frame(
  weekday = ___,
  temperature = ___,
  hot = ___
)
temp
```

```{r exercise-02-01-03-hint-2, eval=FALSE}
temp <- data.frame(
  weekday = c("Monday", "Tuesday", "Wednesday"),
  temperature = ___,
  hot = ___
)
temp
```

```{r exercise-02-01-03-hint-3, eval=FALSE}
temp <- data.frame(
  weekday = c("Monday", "Tuesday", "Wednesday"),
  temperature = c(28, 31, 25),
  hot = ___
)
temp
```

```{r exercise-02-01-03-solution}
temp <- data.frame(
  weekday = c("Monday", "Tuesday", "Wednesday"),
  temperature = c(28, 31, 25),
  hot = c(FALSE, TRUE, FALSE)
)
temp
```

```{r exercise-02-01-03-check}
dftemp <- ex() %>% check_object("temp")
dftemp %>% check_column("weekday") %>% check_equal()
dftemp %>% check_column("temperature") %>% check_equal()
dftemp %>% check_column("hot") %>% check_equal()
ex() %>% check_object("temp") %>% check_equal()
```

## Quiz: Which statements are true about this data frame?

```{r, echo=TRUE, eval=FALSE}
price <- c(28, 31, 25)
data.frame(
  weekday = c("Monday", "Tuesday", "Wednesday", "Thursday"),
  price = price,
  expensive = price > 30
)
```

```{r quiz-02-01-04, echo = FALSE}
question('Which statements are true about the data frame above?',
  answer("The `data.frame()` function will fail because the column `expensive` is no vector.", message = "The column `expensive` is a logical vector resulting from price > 30."),
  answer("The `data.frame()` function will not fail", message = "Check the lengths of the data frame vectors again."),
  answer("The `data.frame()` function fails because the lengths of the vectors are different", correct = TRUE),
   answer('The command would work if `weekday` had the values `c("Monday", "Tuesday", "Wednesday")`', correct = TRUE),
  allow_retry = TRUE
)
```
